PMV: Comparing the prediction accuracy of ISO 7730:2005 and ASHRAE 55:2023

### 2023-09-11 Manuscript/src/sections/results.tex
fix: some pmv values were missing since were outside the applicability limits of the standard

%! Author = sbbfti
%! Date = 10/06/2020
\section{Results and Discussion}\label{sec:results}

\subsection{\ac{db2} overview}\label{subsec:comfort-db-overview}
While the \ac{db2} contains \var{entries_db_all} data points, only \var{entries_db_used} entries met the inclusion criteria listed in the Methodology section.
The distribution of the six input variables used to calculate the PMV indices is visualized in Figure~\ref{fig:dist_input_data} in which we excluded the data points below or above the 2.5th and 97.5th percentiles, respectively.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_input_data}
    \caption{Distribution of the input variables used to calculate the \ac{pmv} and \ac{pmv-ce} values.
    We did not plot data points below or above the 2.5th and 97.5th percentiles, respectively.
    The data are shown using boxen plots (letter-value plots).
    They depict the median as the center-line and each successive level outward contains half of the remaining data.}
    \label{fig:dist_input_data}
\end{figure*}
The input data distributions are far from uniformity (equal density) across the Standards' applicability ranges.
We were not able to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} formulations for inputs that were lower than the 2.5th and 97.5th percentiles, since we only had a limited number of data points beyond those values.
For example, only 2.5\% of the available dataset had a \ac{tdb} $\leq$ than \var{ta_95_perc_min}, this may consequently be an issue since both thermal comfort Standards' applicability limits extend to \qty{10}{\celsius}.
This can be explained by the fact that only a few buildings in the world operate at \ac{tdb} lower than this value.
This is because an indoor \ac{tdb} lower than \var{ta_95_perc_min} would be deemed to be unsatisfactory by a great number of occupants~\cite{iso7730}.
This lack of data in the \ac{db2} is particularly relevant for \ac{v}.
The \ac{pmv} and \ac{pmv-ce} only differ when the value of \ac{v} $\geq$ \qty{0.1}{\m\per\s}, however, the median value in the \ac{db2} was \qty{0.08}{\m\per\s} suggesting that most of the data points were collected in buildings with low air movement.
Consequently, we could not reliably test the accuracy of the \ac{pmv} formulations for values of \ac{v} $\geq$ \var{v_95_perc_max}.

Figure~\ref{fig:dist_other_data} shows the distribution of age, height, weight, and running mean outdoor temperature.
Less than half (\num{23300}) of the total entries have information about the participant's sex and are almost equally distributed among males \qty{52}{\percent} and females.
Ages are not normally distributed, the same is true for the running mean outdoor temperature.
About half of the participants (\qty{52}{\percent}) were aged between \num{20} and \num{35} years old, and only \qty{2.5}{\percent} were older than 60.
This limits our analysis to young adults and does not allow us to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} models in predicting thermal sensation for children or older adults.
The \ac{db2} does not contain information about the health status of the participants.
Approximately \qty{56}{\percent} of the running mean outdoor temperature values where between \qtyrange{10}{25}{\celsius}.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_other_data}
    \caption{Distribution of age, height, weight, and running mean outdoor temperature.
    The data are grouped by sex.
    The text in blue is showing the 2.5th, 50th (median), and 97.5th percentiles.}
    \label{fig:dist_other_data}
\end{figure*}
The percentage of \ac{tpv} grouped by each \ac{tsv} is shown on the left side of Figure~\ref{fig:bar_plot_tp_by_ts}.
Only \var{entries_with_tp} entries had information about both \ac{tsv} and \ac{tpv}.
Thus, on the right, we show a bar chart depicting the distribution of all the \ac{tsv} votes, we analyzed in the paper.
The \ac{tsv} is unbalanced. 
Approximately \var{perc_tsv_neutral} of all the entries have a \ac{tsv} of `Neutral'.
While less than \var{perc_tsv_hot} of the total sample of participants reported to be either `Hot' or `Cold'.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bar_plot_tp_by_ts}
    \caption{The left Figure shows the percentage of \ac{tpv} for each thermal sensation vote.
    The numbers on the right side of each bar show the number of points for each \ac{tsv}.
    The right figure shows the total number of data points grouped by \ac{tsv}.
    Each bin in the right figure has more data points than in the left figure since less data on \ac{tpv} were available.}
    \label{fig:bar_plot_tp_by_ts}
\end{figure*}

In thermal comfort research, it is generally assumed that people who are `slightly warm' or `slightly cool' are thermally comfortable.
However, in the \ac{db2} \qty{68}{\percent} of participants who were `slightly warm' wanted to be `cooler' and \qty{54}{\percent} of participants who were `slightly cool' wanted to be `warmer'.
This finding challenges the above assumption.
Assuming that people who are `slightly warm' or `slightly cool' are comfortable leads to under-counting thermal discomfort in this example.
On the other hand, some participants who reported to be `cool', `cold', `warm', or `hot' wanted `no change'.
Of those who reported to be `cool' \qty{23}{\percent} of them wanted `no change' and \qty{5}{\percent} wanted to be `cooler'.
Similarly, but in reverse, among those who reported to be `warm' \qty{13}{\percent} wanted `no change' and \qty{2}{\percent} wanted to be `warmer'.
This result highlights one of the major limitations of the \ac{tsv} scale and of the assumption that people who are `slightly warm' or `slightly cool' are thermally comfortable.
This suggests that the thermal preference scale is better suited to determine how people perceive their thermal environment.
The \ac{tpv} more clearly allows an HVAC system to decide whether to change or not the thermal environment. 
On the contrary, the \ac{tsv} is ambiguous and does not allow determining if an action should be taken to increase the participant's thermal comfort.

\subsection{Comparison of PMV accuracy in predicting thermal sensation}\label{subsec:model-accuracy-comparison-in-predicting-thermal-sensation}
The \ac{pmv} was developed with the primary aim of predicting \ac{tsv}, consequently, we are here comparing the prediction accuracy of the two formulations.
The two models only differ when \ac{v} exceeds \qty{0.1}{\m\per\s};
hence we are also presenting the results for a subset of the data where \ac{v} is higher than \qty{0.2}{\m\per\s} since above this value the differences in the outputs of the two models are more pronounced.
We set the limit to \qty{0.2}{\m\per\s} and not a higher value since, as previously shown in Figure~\ref{fig:dist_input_data}, we could only reliably test the accuracy of the models up to \ac{v} of \var{v_95_perc_max}.
We did not use \qty{0.1}{\m\per\s} as a threshold because the original difference between the two models was \qty{0.2}{\m\per\s}.
The effect of air movement is perceptible only when the airspeed is sufficiently high to disrupt the body's thermal plume~\cite{zukowska_impact_2012}.
The threshold in the ASHRAE standard was reduced to \qty{0.1}{\m\per\s} to simplify the usability and achieve one threshold for all the air speeds.
% todo:
%  call charts a,b,c,d
\begin{figure*}[htb!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0}
        \caption{All data}
     \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0.2}
        \caption{Only data with \ac{v} $\geq$ \qty{0.2}{\m\per\s}}
     \end{subfigure}
    \caption{The calculated \ac{pmv} value is grouped by the \ac{tsv} value reported by the participants. 
    The number of data points in each bin of the two charts is the same and this allows us to compare the accuracy of both models.
    The overall prediction accuracy is shown above the chart while the model accuracy of predicting participants grouped by their \ac{tsv} is overlayed over each bar. 
    A random model, randomly selecting a thermal sensation vote would have an overall accuracy of \qty{14}{\percent}.}
    \label{fig:bar_stacked_model_accuracy}
\end{figure*}
Being \ac{tsv} the value reported by participants (i.e., ground truth), in Figure~\ref{fig:bar_stacked_model_accuracy} we grouped participants' responses by \ac{tsv} and reported the simple accuracy of the different \ac{pmv} formulations over each sub-figure.
Both \ac{pmv} formulations underestimated the conditions at which participants are thermally dissatisfied with their environment.
The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Being the classification problem unbalanced, the simple accuracy is not an optimal metric to rate the model prediction accuracy since a default strategy of guessing the majority class would lead to a high overall model accuracy.
% todo describe the results for V>0.2
A model that would have always predicted `neutral' would have achieved an overall accuracy of \var{perc_tsv_neutral}, the number of participants who voted \ac{tsv}=0.
For this reason, in Table~\ref{tab:f1}, we report the F1-scores for all the models.
The F1-macro score, which is free from label imbalance, depicts that the accuracy of both models across all metrics is marginally better than random guessing (i.e., \qty{14}{\percent}).
\begin{table}[htb!]
    \centering
    \input{tables/f1.tex}
    \caption{F1-score for the \ac{pmv} and \ac{pmv-ce} models.}
    \label{tab:f1}
\end{table}

% todo describe the table results
The \ac{pmv-ce} was specifically developed to more accurately estimate latent and sensible heat losses from the skin to the environment when air movement is used~\cite{arens_moving_2009}.
Both models had an accuracy lower than \qty{1}{\percent} when predicting either `warm', `hot', or `cold'.
This value is significantly lower and worse than random guessing, this is very concerning because this means that the models are actively misleading the users.
The low prediction accuracy of both model formulations, coupled with the lack of data in the extremes of the applicability limits of the models, may suggest that the models' applicability limits should be revisited.
% todo call it Standards implications, limit use PMV and of environmental variables
Based on our results we recommend restricting the use of the \ac{pmv} only to a values $\lvert \textrm{PMV}\lvert \leq 1.5$.
In addition, we also recommend restricting the ranges on both environmental and personal factors set by both \gls{55} and \gls{7730} Standards.
Currently, these limits extend far beyond the range of data contained in the \ac{db2} dataset.
Being the \ac{db2} the largest thermal comfort dataset in the world and containing data from all continents we deem not to be necessary to have such wide applicability limits for both \gls{55} and \gls{7730} Standards when these environmental variables are rarely observed in buildings across the world.
In addition, the lack of data available beyond the ranges depicted in Figure~\ref{fig:dist_input_data} did not allow us to test the accuracy of both the \ac{pmv} and \ac{pmv-ce} above those ranges.
Consequently, we suggest reducing both Standards applicability limits to the ranges shown in Table~\ref{tab:ranges}, until more data are collected.
\begin{table}[htb!]
    \centering
    \begin{tabular}{cc}
        \toprule
        Variable & Proposed range \\
        \midrule
        \ac{tdb} & \qtyrange{17}{30}{\celsius} \\
        \ac{tr} & \qtyrange{17}{30}{\celsius} \\
        \ac{rh} & \qtyrange{20}{80}{\percent} \\
        \ac{clo} & \qtyrange{0.3}{1.5}{clo} \\
        \ac{met} & \qtyrange{1}{2}{met} \\
        \ac{pmv} & \qtyrange{-1.5}{1.5}{} \\
        % todo shall we also include limits on Age?
        % todo shall we also include airspeed?
        \bottomrule
    \end{tabular}
    \caption{New proposed applicability limits for the \gls{55} and \gls{7730} Standards.}
    \label{tab:ranges}
\end{table}
Grouping the results into discrete categories introduces rounding errors since some participants reported \ac{tsv} on a continuous scale and the \ac{pmv} value is a continuous variable.
To compensate for this we plotted the \ac{pmv} and \ac{pmv-ce} values as a function of \ac{tsv} in Figure~\ref{fig:bubble_models_vs_tsv} and we plotted a \ac{lowess} curve.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bubble_models_vs_tsv}
    \caption{The \ac{lowess} curve shows the relationship between the raw \ac{pmv} and \ac{tsv} data.
    Raw data were then binned and a bubble chart (circle area is proportional to the number of votes in that bin) is superimposed over the regression curve to aid the visualization of a large dataset.}
    \label{fig:bubble_models_vs_tsv}
\end{figure*}
The curve, in the Figure, is calculated using the individual data and not the binned data.
We binned the data to aid the visualization of a large dataset.
If the model is to accurately predict the thermal sensation of people, the regression line should pass through the origin of the cartesian plane and have a slope of 1.
The intercepts for the \ac{pmv} is \qty{-0.23} and for the \ac{pmv-ce} is \qty{-0.24} which means that the model has a bias of less than one-quarter of a thermal sensation interval. 
The intercept is close to 0 and this is aligned with the results shown above that the model can predict people who reported to be neutral \qty{55}{\percent} and \qty{58}{\percent} of the times.
Both \ac{pmv} formulations always underpredicted the thermal sensation of people who reported to be on the extremes of the \ac{tsv}; the slope of the curve was significantly lower than 1.
We discuss in detail in sub-Section~\ref{subsec:sources-of-error} which factors may be the underlying cause of these errors.

@@

\subsection{Model Bias}\label{sec:model-bias}
\subsubsection{Model Overall Bias}\label{subsec:model-overall-bias}
We calculated the overall bias of the model as previously done by \mycite{Humphreys2002} by subtracting the \ac{pmv} and  \ac{pmv-ce} models prediction from the self-reported \ac{tsv}.
Results are presented in Figure~\ref{fig:hist_discrepancies}.
% todo:
%  describe in the paper that we used test to determine that data are not normally distributed.
%  We cannot change into a letter value plot since in that case it would be hard to highlight the number of responses that are within the range from -.5 to.5.
%  The Caption is missing
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/hist_discrepancies}
    \caption{}
    \label{fig:hist_discrepancies}
\end{figure*} 
Overall both the \ac{pmv} and \ac{pmv-ce} models have a median bias lower than half of a thermal sensation interval.
Figure~\ref{fig:hist_discrepancies} contains the summary statistics, and the results highlight that the \ac{pmv-ce} had a higher bias than the \ac{pmv} when using the entire dataset.
The results show that the median bias of the \ac{pmv} model reduced to \var{bias_median_pmv_0.2} from \var{bias_median_pmv_0}, meaning that at higher speed the \ac{pmv} model improved.
Surprisingly, we found that the median bias for the \ac{pmv-ce} increased to \var{bias_median_pmv_ce_0.2} from \var{bias_median_pmv_ce_0}.
These are worrying results since the \ac{pmv-ce} model claims to have a higher accuracy at higher values of \ac{v}, but this is not what was found from this data.
As far as we could find, there is no published paper proving a higher accuracy of the \ac{pmv-ce} compared to the \ac{pmv} model. 
The \ac{pmv-ce} has lower accuracy while making the code more complex and computationally intensive to run.

\subsubsection{Model Bias as Function of Four Input Variables}\label{subsec:model-bias-variable}
In the previous section, we presented the overall bias of the two \ac{pmv} formulations.
We then went on to determine how the bias varied as a function of \ac{tdb}, \ac{v}, \ac{met}, and \ac{clo}.
% todo add humidity ratio chart
We did not include \ac{tr} in the analysis since \ac{tr} was not reported in several studies, hence, we estimated it as detailed in the Methodology.
Furthermore, in many buildings is possible to assume \ac{tr} equal to \ac{tdb} without committing a major error~\cite{Dawe2020}.
Figure~\ref{fig:dist_input_data} shows the results.
In each sub-figure, we have binned the results into discrete intervals of equal width, and the x-axis tick label is the center value of the interval.
All the violin plots have the same area to aid visualization, despite the fact that the number of points in each bin is not constant.
% todo:
%  I need to have the red lines for the same inverse reason why I cannot use the boxenplot in the previous chart

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bias_models}
    \caption{Bias of the \ac{pmv} and \ac{pmv-ce} models plotted as a function of the input variables.}
    \label{fig:bias_models}
\end{figure*}
The results depict that the use of the \ac{pmv-ce} model did not improve the bias of the \ac{pmv} results across all four input variables.
On the contrary, the bias of the \ac{pmv-ce} diverged more from 0 as the value of \ac{v} increased.
These results are in agreement with those presented in the previous section and depict that the correction applied by the \ac{pmv-ce} model worsens the results.

\subsection{Comparison between the PMV formulations with the two-node model}\label{subsec:comparison-between-the-pmv-formulations-wih-the-two-node-model} % todo Stefano Schiavon: call it Gagge PMV model in line with the text below
% todo consider moving to appendix and explain well why I have done this. Stefano Schiavon: The logic and need of this paragraph needs to be better explained. Why did you add? Why can we assume that the PMV-SET is better? Who developed the PMV-SET? We should discuss if we need this paragraph and if it add something that is important and not reported in other sections. In other to do this, we need to explain why we think the PMV-SET is important even if it is not present in standards.

We compared how the results of the \ac{pmv} and \ac{pmv-ce} perform against the results of the two-node model because xxxx. % todo provide information
% todo: Stefano Schiavon: Please provide more details of how these inputs were generated, within which ranges, etc.
Figure~\ref{fig:pmv_two_node_comparison} was generated by calculating the three \ac{pmv} values using the same combination of \num{10000} randomly generated sets inputs.
The \ac{pmv} and \ac{pmv-ce} were then plotted on the y-axis while the \ac{pmvg} value was on the x-axis.
The Figure also shows the \ac{lowess} curves that are used to visualize the prevailing data trend.
% todo Stefano Schiavon: Beside the visualization, can you discuss here also the quantitative results to support the conclusion?
The results of the \ac{pmv} model had a higher agreement than those obtained with the \ac{pmv-ce} model for \ac{pmvg} $\geq$ \qty{0.5}{\m\per\s}. % todo Stefano Schiavon: This is unclear and cannot be deduced from the figure, from where does it come from? How can a PMV model be described by an air speed?
These results show the \ac{pmv-ce} is less similar to the \ac{pmvg} model than the \ac{pmv}, despite using it in the backend to calculate the cooling effect.
This is also an unexpected result and we will discuss why this may be in the following paragraph. 
% todo When a figure report how well a variable is predicted, for example x-measured vs x-predicted, the graph must be a square, otherwise it is visually skewed. 1) I suggest to use the words instead of numbers for PMV 2) We should be consistent with the PMV naming we used (do not use ASHRAE and ISO here) 3) why both models are far from Gagge for PMV<0, both model in general produce a cooler feeling than PMVgagge, we should explain why if we keep this paper.

\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/pmv_two_node_comparison}
    \caption{}
    \label{fig:pmv_two_node_comparison}
\end{figure*}

@@

The \ac{pmv} model uses the overall heat losses or gains, which are only a proxy to the thermal stress that the human body has to undergo to maintain a constant internal temperature, to a \ac{pmv} value which should represent a \ac{tsv} vote.
Fanger created this equation based on a limited dataset collected in a laboratory, and despite the growing body of literature, this equation was never updated or revisited~\cite{Fanger1970}.
% todo Stefano Schiavon: This is vague, Which papers or data is available to update that equation/mapping?
In the \ac{pmv} model the heat losses and gains are mapped using constant coefficients to a thermal sensation vote, disregarding the fact that the human body employs different strategies, i.e., vasodilation and vasoconstriction to cope with warm and cold conditions, respectively~\cite{romanovsky_thermoregulation_2018}.

>>>>

\section{Results and Discussion}\label{sec:results}

\subsection{\ac{db2} overview}\label{subsec:comfort-db-overview}
While the \ac{db2} contains \var{entries_db_all} data points, only \var{entries_db_used} entries met the inclusion criteria listed in the Methodology section.
The distribution of the six input variables used to calculate the PMV indices is visualized in Figure~\ref{fig:dist_input_data}.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_input_data}
    \caption{Distribution of the input variables used to calculate the \ac{pmv} and \ac{pmv-ce} values.
    The data are shown using boxen plots (letter-value plots).
    They depict the median as the center-line and each successive level outward contains half of the remaining data.}
    \label{fig:dist_input_data}
\end{figure*}
The input data distributions are far from uniformity (equal density) across the Standards' applicability ranges.
We were not able to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} formulations for inputs that were lower than the 2.5th and 97.5th percentiles, since we only had a limited number of data points beyond those values.
For example, only \qty{2.5}{\percent} of the available dataset had a \ac{tdb}~$\leq$ than \var{ta_95_perc_min}, this is a problem since both thermal comfort Standards' applicability limits extend to \qty{10}{\celsius}.
This can be explained by the fact that only a few buildings in the world operate at \ac{tdb} lower than this value.
Mainly because an indoor \ac{tdb} lower than \var{ta_95_perc_min} would be deemed to be unsatisfactory by a great number of occupants~\cite{iso7730}.
This lack of data in the \ac{db2} is particularly relevant for \ac{v}.
The \ac{pmv} and \ac{pmv-ce} only differ when the value of \ac{v}~$\geq$~\qty{0.1}{\m\per\s}, however, the median value in the \ac{db2} was \qty{0.08}{\m\per\s} suggesting that most of the data points were collected in buildings with low air movement.
Consequently, we could not reliably test the accuracy of the \ac{pmv} formulations for values of \ac{v}~$\geq$~\var{v_95_perc_max}.
% todo v_95_perc_max these variables should be calculated dropping nan

Figure~\ref{fig:dist_other_data} shows the distribution of age, height, weight, and running mean outdoor temperature.
Less than half (\num{23300}) of the total entries have information about the participant's sex and are almost equally distributed among males \qty{52}{\percent} and females.
Ages are not normally distributed, the same is true for the running mean outdoor temperature.
About half of the participants (\qty{52}{\percent}) were aged between \num{20} and \num{35} years old, and only \qty{2.5}{\percent} were older than 60.
This limits our analysis to young adults and does not allow us to determine the accuracy of the \ac{pmv} and \ac{pmv-ce} models in predicting thermal sensation for children or older adults.
The \ac{db2} does not contain information about the health status of the participants.
Approximately \qty{56}{\percent} of the running mean outdoor temperature values where between \qtyrange{10}{25}{\celsius}.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/dist_other_data}
    \caption{Distribution of age, height, weight, and running mean outdoor temperature.
    The data are grouped by sex.
    The text in blue is showing the 2.5th, 50th (median), and 97.5th percentiles.}
    \label{fig:dist_other_data}
\end{figure*}

The percentage of \ac{tpv} grouped by each \ac{tsv} is shown on the left side of Figure~\ref{fig:bar_plot_tp_by_ts}.
Only \var{entries_with_tp} entries had information about both \ac{tsv} and \ac{tpv}.
Thus, on the right, we show a bar chart depicting the distribution of all the \ac{tsv} votes, we analyzed in the paper.
The \ac{tsv} dataset is unbalanced.
Approximately \var{perc_tsv_neutral} of all the entries have a \ac{tsv} of `Neutral'.
While less than \var{perc_tsv_hot} of the total sample of participants reported to be either `Hot' or `Cold'.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bar_plot_tp_by_ts}
    \caption{The left Figure shows the percentage of \ac{tpv} for each thermal sensation vote.
    The numbers on the right side of each bar show the number of points for each \ac{tsv}.
    The right figure shows the total number of data points grouped by \ac{tsv}.
    Each bin in the right figure has more data points than in the left figure since less data on \ac{tpv} were available.}
    \label{fig:bar_plot_tp_by_ts}
\end{figure*}

In thermal comfort research, it is generally assumed that people who are `slightly warm' or `slightly cool' are thermally comfortable.
However, in the \ac{db2} \qty{68}{\percent} of participants who were `slightly warm' wanted to be `cooler' and \qty{54}{\percent} of participants who were `slightly cool' wanted to be `warmer'.
This finding challenges the above assumption.
Assuming that people who are `slightly warm' or `slightly cool' are comfortable leads to under-counting thermal discomfort in our dataset.
On the other hand, some participants who reported to be `cool', `cold', `warm', or `hot' wanted `no change'.
Of those who reported to be `cool' \qty{23}{\percent} of them wanted `no change' and \qty{5}{\percent} wanted to be `cooler'.
Similarly, but in reverse, among those who reported to be `warm' \qty{13}{\percent} wanted `no change' and \qty{2}{\percent} wanted to be `warmer'.
This result highlights one of the major limitations of the \ac{tsv} scale and of the assumption that people who are `slightly warm' or `slightly cool' are thermally comfortable.
This suggests that the thermal preference scale is better suited to determine how people perceive their thermal environment.
The \ac{tpv} more clearly allows an HVAC system to decide whether to change or not the thermal environment. 
On the contrary, the \ac{tsv} is ambiguous and does not allow determining if an action should be taken to increase the participant's thermal comfort.

\subsection{Comparison of PMV accuracy in predicting thermal sensation}\label{subsec:model-accuracy-comparison-in-predicting-thermal-sensation}
The \ac{pmv} was developed with the primary aim of predicting \ac{tsv}, consequently, we are comparing the predicted value from the two models with measured \ac{tsv}.
The two models only differ when \ac{v} exceeds \qty{0.1}{\m\per\s};
hence we also present the results for a subset of the data where \ac{v} is higher than \qty{0.2}{\m\per\s} since above this value the differences in the outputs of the two models are more pronounced.
We set the limit to \qty{0.2}{\m\per\s} and not a higher value since, as previously shown in Figure~\ref{fig:dist_input_data}, we could only reliably test the accuracy of the models up to \ac{v} of \var{v_95_perc_max}.
We did not use \qty{0.1}{\m\per\s} as a threshold because the original difference between the two models was \qty{0.2}{\m\per\s} and it was change to \qty{0.1}{\m\per\s} only in the \gls{55}.
The effect of air movement is perceptible only when the airspeed is sufficiently high to disrupt the body's thermal plume~\cite{zukowska_impact_2012}.
The threshold in the ASHRAE standard was reduced to \qty{0.1}{\m\per\s} to simplify the usability and achieve one threshold for all the air speeds.
% todo call charts a,b,c,d
\begin{figure*}[htb!]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0}
        \caption{All data}
     \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/bar_stacked_model_accuracy_0.2}
        \caption{Only data with \ac{v}~$\geq$~\qty{0.2}{\m\per\s}}
     \end{subfigure}
    \caption{The calculated \ac{pmv} value is grouped by the \ac{tsv} value reported by the participants. 
    The number of data points in each bin of the two charts is the same and this allows us to compare the accuracy of both models.
    The overall prediction accuracy is shown above the chart while the model accuracy of predicting participants grouped by their \ac{tsv} is overlayed over each bar. 
    A random model, randomly selecting a thermal sensation vote would have an overall accuracy of \qty{14}{\percent}.}
    \label{fig:bar_stacked_model_accuracy}
\end{figure*}
Being \ac{tsv} the value reported by participants (i.e., ground truth), in Figure~\ref{fig:bar_stacked_model_accuracy} we grouped participants' responses by \ac{tsv} and reported the simple accuracy of the different \ac{pmv} formulations over each sub-figure.
Both \ac{pmv} formulations underestimated the conditions at which participants are thermally dissatisfied with their environment.
The overall accuracy of the \ac{pmv} and \ac{pmv-ce} were \qty{33}{\percent} and \qty{34}{\percent}, respectively.
Being the classification problem unbalanced, the simple accuracy is not an optimal metric to rate the model prediction accuracy since a default strategy of guessing the majority class would lead to a high overall model accuracy.
A model that would have always predicted `neutral' would have achieved an overall accuracy of \var{perc_tsv_neutral}, the number of participants who voted \ac{tsv}~=~0.
For both models the simple accuracy decreases when we only include entries with \ac{v}~$\geq$~\qty{0.2}{\m\per\s}.
We want to remind the reader that the number of datapoints in each bin is different.

In Table~\ref{tab:f1}, we report the F1-scores for both models, using the whole dataset, only keeping entries with \ac{v}~$\geq$~\qty{0.2}{\m\per\s}, and for those with a $\lvert \textrm{PMV}\lvert \leq 1.5$ and $\lvert \textrm{TSV}\lvert \leq 1.5$.
The F1-macro score, which is free from label imbalance, depicts that the accuracy of both models across all metrics is marginally better than random guessing (i.e., \qty{14}{\percent}) for all the dataset.
For the subset of data with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} the F1-scores of the \ac{pmv-ce} are at best \num{0.01} point higher than \ac{pmv} which does not justify its superiority.
On the contrary the F1-macro score of the \ac{pmv} is \num{.23}, \qty{44}{\percent} higher than the \ac{pmv-ce} which is \num{.16} for subset of data with $\lvert \textrm{PMV}\lvert \leq 1.5$ and $\lvert \textrm{TSV}\lvert \leq 1.5$.
These results imply that the \ac{pmv} is more accurate than the \ac{pmv-ce} within the range in which both model perform better, i.e., closer to thermo neutrality.
\begin{table}[htb!]
    \centering
    \input{tables/f1.tex}
    \caption{F1-score for the \ac{pmv} and \ac{pmv-ce} models for three datasets.}
    \label{tab:f1}
\end{table}

Grouping the results into discrete categories introduces rounding errors since some participants reported \ac{tsv} on a continuous scale and the \ac{pmv} value is a continuous variable.
To compensate for this we plotted the \ac{pmv} and \ac{pmv-ce} values as a function of \ac{tsv} in Figure~\ref{fig:bubble_models_vs_tsv} and we plotted a \ac{lowess} curve.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bubble_models_vs_tsv}
    \caption{The \ac{lowess} curve shows the relationship between the raw \ac{pmv} and \ac{tsv} data.
    Raw data were then binned and a bubble chart (circle area is proportional to the number of votes in that bin) is superimposed over the regression curve to aid the visualization of a large dataset.}
    \label{fig:bubble_models_vs_tsv}
\end{figure*}
The curve, in the Figure, is calculated using the individual data and not the binned data.
We only binned the data to aid the visualization of a large dataset.
If the model is to accurately predict the thermal sensation of people, the regression line should pass through the origin of the cartesian plane and have a slope of 1.
The intercepts for the \ac{pmv} is \qty{-0.23} and for the \ac{pmv-ce} is \qty{-0.24} which means that the model has a bias of less than one-quarter of a thermal sensation interval. 
This result is aligned with those presented in Figure~\ref{fig:bar_stacked_model_accuracy} where the \ac{pmv} and \ac{pmv-ce} models can predict people who reported to be neutral \qty{55}{\percent} and \qty{58}{\percent} of the times, respectively.
Both \ac{pmv} formulations always under-predicted the thermal sensation of people who reported to be on the extremes of the \ac{tsv}; the slope of the curve was significantly lower than 1.
We discuss in detail in sub-Section~\ref{subsec:sources-of-error} which factors may be the underlying cause of these errors.

@@

\subsection{Model Bias}\label{subsec:model-bias}
\subsubsection{Model Overall Bias}\label{subsubsec:model-overall-bias}
We calculated the overall bias of the model as previously done by \mycite{Humphreys2002} by subtracting the \ac{pmv} and  \ac{pmv-ce} models prediction from the self-reported \ac{tsv}.
Results are presented in Figure~\ref{fig:hist_discrepancies}.
%  We cannot change into a letter value plot since in that case it would be hard to highlight the number of responses that are within the range from -.5 to.5.
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/hist_discrepancies}
    \caption{}
    \label{fig:hist_discrepancies}
\end{figure*} 
Overall both the \ac{pmv} and \ac{pmv-ce} models have a median bias lower than half of a thermal sensation interval.
Figure~\ref{fig:hist_discrepancies} contains the summary statistics, and the results highlight that the \ac{pmv-ce} had a higher bias than the \ac{pmv} when using the entire dataset.
When including only the data with \ac{v}~$\geq$~\qty{0.2}{\m\per\s} the median bias of the \ac{pmv} model reduced to \var{bias_median_pmv_0.2} from \var{bias_median_pmv_0}, meaning that at higher speed the \ac{pmv} model improved.
Surprisingly, we found that the median bias for the \ac{pmv-ce} increased to \var{bias_median_pmv_ce_0.2} from \var{bias_median_pmv_ce_0}.
These are worrying results since the \ac{pmv-ce} model claims to have a higher accuracy at higher values of \ac{v}, but this is not what was found from this data.
The \ac{pmv-ce} has a higher bias while making the code more complex and computationally intensive to run.

\subsubsection{Model Bias as Function of Five Input Variables}\label{subsubsec:model-bias-variable}
In the previous section, we presented the overall bias of the two \ac{pmv} formulations.
We then went on to determine how the bias varied as a function of \ac{tdb}, \ac{v}, \ac{rh}, \ac{met}, and \ac{clo}.
We did not include \ac{tr} in the analysis since \ac{tr} was not measured in several studies, hence, we estimated it as detailed in the Methodology.
Furthermore, in many buildings is possible to assume \ac{tr} equal to \ac{tdb} without committing a major error~\cite{Dawe2020}.
Figure~\ref{fig:dist_input_data} shows the results.
In each sub-figure, we have binned the results into discrete intervals of equal width, and the x-axis tick label is the center value of the interval.
All the violin plots have the same area to aid visualization, despite the fact that the number of points in each bin is not constant.
%  I need to have the red lines for the same inverse reason why I cannot use the boxenplot in the previous chart
% todo in the plot do not report the mean value but the range instead since for example V max is 0.46m/s
% todo report p-values
\begin{figure*}[htb!]
    \centering
    \includegraphics[width=\textwidth]{figures/bias_models}
    \caption{Bias of the \ac{pmv} and \ac{pmv-ce} models plotted as a function of the input variables.}
    \label{fig:bias_models}
\end{figure*}
The results depict that the use of the \ac{pmv-ce} model did not improve the bias of the \ac{pmv} results across all four input variables.
On the contrary, the bias of the \ac{pmv-ce} diverged more from 0 as the value of \ac{v} increased.
These results are in agreement with those presented in the previous section and depict that the correction applied by the \ac{pmv-ce} model worsens the results.
Both model had a median bias higher or equal to $\lvert0.5\lvert$ for \ac{tdb}~$\leq$~\qty{20}{\celsius}; \ac{clo}~$\leq$~\qty{.6}{clo} and $\geq$~\qty{1.2}{clo}.
Surprisingly again the bias of the \ac{pmv-ce} model was significantly worse than the one of the \ac{pmv} for \ac{v}~$\geq$~\qty{0.2}{\m\per\s}.
% todo report the p-value

@@

The \ac{pmv} model uses the overall heat losses or gains to calculate a \ac{pmv} value which should represent a \ac{tsv} vote.
Heat losses or gains are only a proxy to the thermal stress that the human body has to undergo to maintain a constant internal temperature.
Fanger created this equation based on a limited dataset collected in a laboratory, and despite the growing body of literature, this equation was never updated or revisited~\cite{Fanger1970}.
% todo Stefano Schiavon: This is vague, Which papers or data is available to update that equation/mapping?
In the \ac{pmv} model the heat losses and gains are mapped using constant coefficients to a thermal sensation vote, disregarding the fact that the human body employs different strategies, i.e., vasodilation and vasoconstriction to cope with warm and cold conditions, respectively~\cite{romanovsky_thermoregulation_2018}.
